---
title: "Vegas"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Two-arm trial TFP v RStan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

## Quick start
The model implemented here is a non-centered parameterization for a logistic model.


```{r rsetup, include=FALSE,eval=TRUE}
library(reticulate)
use_virtualenv("/Users/gogol/py_env/bayes", required = TRUE)
```

## Data set

The R chunk below creates a simple dataset of three cols:

-   a binary response variable (0/1 = non-responder/responder),
-   a binary treatment variable (0/1 = control/test treatment)
-   and a basket ID variable. (1)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. a classical two arm randomized trial design. Baskets will be added in other vignettes.

```{r datasets, include=TRUE,eval=TRUE}
### Prepare model inputs
set.seed(99999)
# Set up data
rr_k_ctrl <- c(0.20)        # control response rate for each basket
rr_k_trt <- c(0.40)         # treatment response rate for each basket

K<-length(rr_k_ctrl)        # number of baskets

N_k_ctrl <- rep(1500, K)     # number of control participants per basket
N_k_trt <- rep(1500, K)      # number of treatment participants per basket
N_k <- N_k_ctrl + N_k_trt   # number of participants per basket (both arms combined)
N <- sum(N_k)               # total sample size
k_vec <- rep(1:K, N_k)      # N x 1 vector of basket indicators (1 to K)

T_vec<-NULL;
y<-NULL;
for(i in 1:K){ # for each basket repeat 0-control 1-trt according to the specifc Ns
  T_vec<-c(T_vec,rep(0:1,c(N_k_ctrl[i],N_k_trt[i]))) # treatment/control indicator
  y<-c(y,
       c(rbinom(N_k_ctrl[i],1,rr_k_ctrl[i]), # bernoulli for control
         rbinom(N_k_trt[i],1,rr_k_trt[i]))) #           for trt
}

thedata<-data.frame(y,basketID=k_vec,Treatment=T_vec)

py$thedata<-r_to_py(thedata) # THE KEY LINE - makes data available to Python
                             # this is converted into a pandas dataframe object in Python

eta=0.1+0.5*T_vec[1:10]
p=exp(eta)/(1+exp(eta))

sum(y[1:10]*log(p)+(1-y[1:10])*log(1-p))

eta=-1.1+1.5*T_vec[1:10]
p=exp(eta)/(1+exp(eta))

exp(sum(y[1:10]*log(p)+(1-y[1:10])*log(1-p)))

summary(m1<-glm(y~1+Treatment,data=thedata[,],family="binomial"))

summary(m0<-glm(y~1,data=thedata[,],family="binomial"))

```

### Setup
```{python onevar,include=TRUE,eval=TRUE}
import numpy as np
import pandas as pd
#from jax import config
#config.update("jax_enable_x64", True)
import jax.numpy as jnp
import jax
import vegas
import math
import gvar

## The data.frame passed is now a panda df but for TPF this needs to be further
## converted into tensors here we simply convert each col of the data.frame into
## a Rank 1 tensor (i.e. a vector)
y_data=jnp.array(thedata.iloc[:,0],dtype=jnp.float32)[:,None] # 1 col matrix
k_vec=jnp.array(thedata.iloc[:,1])-1
T_vec=jnp.array(thedata.iloc[:,2],dtype=jnp.float32)[:,None] # 1 col matrix


pars=jnp.array([[0.1,-1.1,2.1]],dtype=jnp.float32)
#pars=jnp.stack([pars, pars,pars], axis=1)


@jax.jit
def compute_log_lik(theta, y,T):
    #print(f"shape1={x.shape}\n")
    """
    logL = sum_i^n ( y_i*(a0+a1*T_i) - log(1+ exp(a0+a1*T_i) )  )
    Expects x to have shape (batch_size, DIM).
    """
    #print(f"shape={theta.shape}\n")
    a0=theta[0,:]#,:] # (3,)
    eta=a0# (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - jnp.log(1+jnp.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)),axis=0) #
    # now add the priors
    # simple N(0,sd=10)
    prior_a0=jax.scipy.stats.norm.logpdf(a0,loc=0.,scale=10.)
    logDens = logL + prior_a0 
    return logDens # this will be (3,)

# 2. Setup Vegas with a Shifting Logic
@vegas.rbatchintegrand
class StableLikelihood:
    def __init__(self, y, T):
        self.y = y
        self.T = T
        self.l_max = 0.0 # Will be updated after warm-up

    def __call__(self, theta):
        theta_jax = jnp.array(theta)
        log_lik = compute_log_lik(theta_jax, self.y,self.T)
        
        # Shift by the maximum found so far
        # This keeps the result near 1.0 instead of 1e-300
        return jnp.exp(log_lik - self.l_max)

model = StableLikelihood(y=y_data,T=T_vec)

start = jnp.array([-2],dtype=jnp.float32) # a0 a1
stop = jnp.array([1],dtype=jnp.float32)   # a0 a1
searchPts=jnp.transpose(jnp.linspace(start, stop, 500))

model.l_max=jnp.max(compute_log_lik(searchPts, model.y,model.T))

#explicitly set seed
gvar.ranseed(9999)

integ = vegas.Integrator(jnp.array([[-2, 1]]*1,dtype=jnp.float32))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=10000)
# STAGE 2: Production (Stable Integration)
result = integ(model, nitn=10, neval=50000)

# 5. Output results
print(f"\nIntegration Results:")
print(f"Result: {result.mean:.8f} +/- {result.sdev:.8f}")
print(f"Q-value: {result.Q:.2f} (should be close to 0.5 for a good fit)")

# STAGE 3: Final Log-Evidence Calculation
# log(I) = l_max + log(scaled_integral)
log_evidence = model.l_max + jnp.log(result.mean)

print(f"Log Evidence: {log_evidence}")

# we want posterior values
# compute_log_lik(theta, y,T) - log_evidence = log( f(D|a0)f(a0) )

mydens=np.array(jnp.exp(compute_log_lik(searchPts,y_data,T_vec) - log_evidence))
myX=np.array(searchPts)

```

Now let us try to unpack what is going on in the above code.

## Code Explanation: Bayesian Logistic Regression with VEGAS Integration

This code performs Bayesian inference for a logistic regression model using JAX and the VEGAS Monte Carlo integration library. It estimates the marginal likelihood (evidence) for parameter $a_0$ by integrating the unnormalized posterior over its prior distribution.

### 1. Data Preparation
```python
y_data=jnp.array(thedata.iloc[:,0],dtype=jnp.float32)[:,None]  # Response variable (n×1)
k_vec=jnp.array(thedata.iloc[:,1])-1                           # Index variable (not used here)
T_vec=jnp.array(thedata.iloc[:,2],dtype=jnp.float32)[:,None]   # Predictor variable (n×1)
```
Converts pandas DataFrame columns to JAX tensors. `y_data` contains binary outcomes $y_i \in \{0,1\}$, and `T_vec` contains binary treatment predictor values $T_i$.

### 2. Model Parameters and Log-Likelihood
```python
pars=jnp.array([[0.1,-1.1,2.1]],dtype=jnp.float32)
```
Initial parameter guess (not used in final computation).

**Log-likelihood function:**
```python
@jax.jit
def compute_log_lik(theta, y, T):
    a0 = theta[0,:]                    # Extract parameter a₀
    eta = a0                           # Linear predictor: ηᵢ = a₀ (broadcast across observations)
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)), axis=0)  # Log-likelihood sum
    prior_a0 = jax.scipy.stats.norm.logpdf(a0, loc=0., scale=10.)  # N(0,10) prior
    logDens = logL + prior_a0          # Unnormalized log-posterior
    return logDens
```

The above JAX function defines the log likelihood for a Benoulli random variable $y_i \sim Bernoulli(p_i)$, with

$$
p_i = \frac{1}{1 + e^{-\eta_i}}, where \quad \eta_i = a_0 + a_1T_i 
$$

In this case, we are implementing an intercept only model, i.e., $\eta_i = a_0$. Therefore, 
$$
P(y_i=1|T_i) = \frac{1}{1 + e^{-(a_0)}}
$$
The log-likelihood for a single observation is given by

$$
\log p(yi|a_0) = y_i\eta_i - log(1 + e^{\eta_i})
$$
And for \(n\) observations is given by:

$$
\log L(a_0) = \sum_{i=1}^n \left[ y_i \cdot \eta_i - \log(1 + e^{\eta_i}) \right]
$$
Since $\eta_i = a_0$ in this cse, this simplifies to
$$
\log L(a_0) = \sum_{i=1}^n \left[ y_i \cdot a_0 - \log(1 + e^{a_0}) \right]
$$
The prior for $a_0$ is drawn as $a_0 \sim N(mean =0, variance = 10^2).$ using the ```jax.scipy.stats.norm.logpdf``` function. Since this prior is in log scale, we get,

$$
\log p(a_0) = -1/2(a_0/10)^2 - \log (10(2\pi)^{-1/2}) \quad From \quad Normal \quad pdf
$$

Now $Posterior \: density \propto Likelihood X Prior$. Therefore on the log scale, $\log (Posterior \: density) \propto \log (Likelihood) + \log (Prior)$ 

Which gives is the unnormalized posterior (numerator) as:
$$
\log f(a_0) = \log L(a_0) + \log p(a_0)
$$


No that we have our numerator, let us see how we can find the denominator, i.e., the evidence or marginal, so that we can get the normalized desnity. Here is where we can see the full power of ```vegas```.

### 3. VEGAS Integration Setup
```python
@vegas.rbatchintegrand
class StableLikelihood:
    def __init__(self, y, T):
        self.y = y
        self.T = T
        self.l_max = 0.0  # Maximum log-likelihood (for numerical stability)

    def __call__(self, theta):
        theta_jax = jnp.array(theta)
        log_lik = compute_log_lik(theta_jax, self.y, self.T)
        return jnp.exp(log_lik - self.l_max)  # Shifted integrand
        
model = StableLikelihood(y=y_data,T=T_vec)
```

But before that, **What is shifting logic?** 

In Bayesian computation, we often need to evaluate $f(\theta) = exp(log(f(\theta))$. Where, $\log(f(\theta) = \log p(y | \theta) + \log p(\theta)$.

However, $\log(f(\theta))$ is often a large negative number, e.g., -1000 or -2000, making $f(\theta) = exp(log(f(\theta)) = exp(-1000) \approx 0$. This causes numerical underflow, meaning the computer treats values as zero, which breaks Monte Carlo integration.

**The core idea of shifting**

To avoid this, we substact a constant $l_{max} = \underset{\theta}{max} \log f(\theta)$. Then redefine $f^* = exp(\log f(\theta) - l_{max})$. Since $\log f(\theta) - l_{max}\le 0$, we get, $\lt f^* \le 1$.

Interestingly, this hack can be very easily taken care of while performing the integration as follows:

We want, $I = \int f(\theta)d\theta$

Using the shifting function, we get $I = \int exp(\log f(\theta))d\theta$

Writing $\log f(\theta)$ as $\log f(\theta) = (\log f(\theta) - l_{max}) + l_{max}$ and substitutuing it for the above integrand, we get $I = \int exp(\log f(\theta) - l_{max}).exp(l_{max})d\theta = exp(l_{max})\cdot\int f^*(\theta)d\theta$

This is exactly what the above code chunk is doing. It is initializing $l_{max}$ with $0$, computing the log likelihood for $\log f(\theta)$ using the `JAX` `compute_log_lik` function we had defined earlier, and returning the shifted integrand $\int f^*(\theta)d\theta = \int exp(\log f(\theta) - l_{max})d\theta$. During the wamup $l_max$ is updated with the meximum found so far. 


### 4. Integration Process
```python
start = jnp.array([-2], dtype=jnp.float32)
stop = jnp.array([1], dtype=jnp.float32)
searchPts = jnp.transpose(jnp.linspace(start, stop, 500))
model.l_max = jnp.max(compute_log_lik(searchPts, model.y, model.T))
```

This part is doing grid search for the Monte Carlo sampling with

- Creates 500 grid points from \([-2, 1]\) to find maximum log-posterior
- Updates `l_max`

```python
integ = vegas.Integrator(jnp.array([[-2, 1]]*1,dtype=jnp.float32))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=10000)
# STAGE 2: Production (Stable Integration)
result = integ(model, nitn=10, neval=50000)
```

**Two-stage VEGAS integration:**
1. **Warm-up** (`nitn=10, neval=10000`): Adapts integration grid
2. **Production** (`nitn=10, neval=50000`): Final estimate

### 5. Evidence Calculation
```python
log_evidence = model.l_max + jnp.log(result.mean)
```
**Final result:** $\log I = l_{\max} + \log(\hat{I}))$, where $\hat{I}$ is the mean of VEGAS estimate of the shifted integral.

**Q-value interpretation:** Should be ~0.5 for well-converged integration.

### 6. Posterior Samples
```python
mydens = np.array(jnp.exp(compute_log_lik(searchPts, y_data, T_vec) - log_evidence))
```
Computes the normalized **posterior densities** on grid:
$$
p(a_0|D) = \frac{p(D|a_0) \pi(a_0)}{I}, \quad I = \exp(\log I)
$$

Or in the log scale,

$$
\log p(a0∣y)=logf(a_0)−logI \implies p(a0∣y)=exp(f(a0)−logI)
$$

We estimated $f(a_0)$ in step 2 and $I$ in step 3. The above code replaces this values to find the log posterior.

**Summary:** 

1. Convert data into tensors

2. Define log-likelihood + prior

3. Compute unnormalized log posterior

4. Define the integrand for evidence

5. Estimate the evidence (marginal)

5. Compute the normalized log posterior desnity


**Now we are ready to plot our posterior**
```{r onevarR,include=TRUE,eval=TRUE}

plot(py$myX,py$mydens,type="l")
abline(v=coef(m0))

```


```{python onevar_withjaco,include=TRUE,eval=TRUE}
import numpy as np
import pandas as pd
import jax.numpy as jnp
import jax
import vegas
import math
import gvar

y_data=jnp.array(thedata.iloc[:,0],dtype=jnp.float32)[:,None] # 1 col matrix
k_vec=jnp.array(thedata.iloc[:,1])-1
T_vec=jnp.array(thedata.iloc[:,2],dtype=jnp.float32)[:,None] # 1 col matrix


pars=jnp.array([[-0.1,0.2,0.3],[0.01,-0.9,0.05]],dtype=jnp.float32)
#pars=jnp.stack([pars, pars,pars], axis=1)

#change of variables, we take a0 and a1 both on -infty to + infty
# to a0 = t1/(1-t1^2) and a1 = t2/(1-t2^2)
# and jacobian is a = (1+t^2)/(1-t^2)^2
# so evaluate function at these transformations and add in jacobian terms


@jax.jit
def compute_log_lik_adj(theta, y,T):
    #print(f"shape1={x.shape}\n")
    """
    logL = sum_i^n ( y_i*(a0+a1*T_i) - log(1+ exp(a0+a1*T_i) )  )
    Expects x to have shape (batch_size, DIM).
    """
    #print(f"shape={theta.shape}\n")
    #a0=theta[0,:]#,:] # (3,)
    #a1=theta[1,:]#,:] # (3,)
    
    theta = jnp.clip(theta, -0.99999, 0.99999) 
    
    a0=theta[0,:]/(1-jnp.power(theta[0,:],2))
    
    jacobianL = jnp.log(1+jnp.power(theta[0,:],2)) - 2.0*jnp.log(1-jnp.power(theta[0,:],2))
    
    eta=a0 # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - jnp.log(1+jnp.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)),axis=0) #
    # now add the priors
    # simple N(0,sd=10)
    prior_a0=jax.scipy.stats.norm.logpdf(a0,loc=0.,scale=10.)
    logDens = logL + prior_a0
    logL = logDens+jacobianL
    return logL # this will be (3,)


# 2. Setup Vegas with a Shifting Logic
@vegas.rbatchintegrand
class StableLikelihood:
    def __init__(self, y, T):
        self.y = y
        self.T = T
        self.l_max = 0.0 # Will be updated after warm-up

    def __call__(self, theta):
        theta_jax = jnp.array(theta)
        log_lik = compute_log_lik_adj(theta_jax, self.y,self.T)
        
        # Shift by the maximum found so far
        # This keeps the result near 1.0 instead of 1e-300
        return jnp.exp(log_lik - self.l_max)

model = StableLikelihood(y=y_data,T=T_vec)

start = jnp.array([-1],dtype=jnp.float32) # a0 a1
stop = jnp.array([1],dtype=jnp.float32)   # a0 a1
searchPts=jnp.transpose(jnp.linspace(start, stop, 500,endpoint=False))

model.l_max=jnp.max(compute_log_lik_adj(searchPts, model.y,model.T))
#explicitly set seed
gvar.ranseed(9999)

integ = vegas.Integrator(jnp.array([[-1, 1]]*1,dtype=jnp.float32))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=1000)
# STAGE 2: Production (Stable Integration)
result = integ(model, nitn=10, neval=5000)

# 5. Output results
print(f"\nIntegration Results:")
print(f"Result: {result.mean:.8f} +/- {result.sdev:.8f}")
print(f"Q-value: {result.Q:.2f} (should be close to 0.5 for a good fit)")

# STAGE 3: Final Log-Evidence Calculation
# log(I) = l_max + log(scaled_integral)
log_evidence = model.l_max + jnp.log(result.mean)

print(f"Log Evidence: {log_evidence}")


start2 = jnp.array([-0.7],dtype=jnp.float32) # a0 a1
stop2 = jnp.array([0.2],dtype=jnp.float32)   # a0 a1
searchPts2=jnp.transpose(jnp.linspace(start2, stop2, 500,endpoint=False))


# to map from -1 +1 mapping to -inf + inf - searchPts2 on -1 +1
def jac_trans(theta):
    theta = jnp.clip(theta, -0.99999, 0.99999) 
    a0=theta[0,:]/(1-jnp.power(theta[0,:],2))
    return(a0)

# need to turn off jacobian in density - just need density after mapped from -1 +1 to -inf +inf
def compute_log_lik_adj2(theta, y,T):
    
    theta = jnp.clip(theta, -0.99999, 0.99999) 
    
    a0=theta[0,:]/(1-jnp.power(theta[0,:],2))
    
    eta=a0 # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - jnp.log(1+jnp.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)),axis=0) #
    # now add the priors
    # simple N(0,sd=10)
    prior_a0=jax.scipy.stats.norm.logpdf(a0,loc=0.,scale=10.)
    logDens = logL + prior_a0
    logL = logDens
    return logL # this will be (3,)

    
# we want posterior values
# compute_log_lik(theta, y,T) - log_evidence = log( f(D|a0,a1)f(a0)f(a1) )
mydens2=np.array(jnp.exp(compute_log_lik_adj2(searchPts2,y_data,T_vec) - log_evidence))
myX2=np.array(jac_trans(searchPts2))


```



# Explanation of the Code (With Change of Variables)

## 1. Overview

This code performs **Bayesian inference for a Bernoulli/logistic model**, but introduces an important modification:

> A **change of variables** to transform an **unbounded parameter space** into a **bounded domain** for numerical integration.

The goal remains:

* Compute **model evidence**
* Compute **posterior density**

---

## 2. Original Parameter Space

The parameter is:

$$
a_0 \in (-\infty, +\infty)
$$

This creates a challenge for numerical integration because:

* VEGAS requires **finite bounds**
* Infinite domains are difficult to sample efficiently

---

## 3. Change of Variables

To address this, the code introduces:

$$
a_0 = \frac{\theta}{1 - \theta^2}, \quad t \in (-1,1)
$$

This mapping has the property:

$$
\theta \in (-1,1) \quad \Longleftrightarrow \quad a_0 \in (-\infty, +\infty)
$$

---

## 4. Jacobian of the Transformation

When changing variables, we must adjust the density:

$$
\int f(a_0), da_0 = \int f(a_0(\theta)) \left| \frac{da_0}{d\theta} \right| d\theta
$$

The derivative is:

$$
\frac{da_0}{d\theta} = \frac{1 + \theta^2}{(1 - \theta^2)^2}
$$

So the log-Jacobian is:

$$
\log \left| \frac{da_0}{d\theta} \right| = \log(1+\theta^2) - 2\log(1-\theta^2)
$$

This is exactly what the following code in the function `compute_log_lik_adj` computes:

```python
jacobianL = log(1+t^2) - 2 log(1-t^2)
```

---

## 5. Adjusted Log Density

The transformed log-density then becomes:

$$
\log f^*(\theta) = \log f(a_0(\theta)) + \log \left| \frac{da_0}{d\theta} \right|
$$

This ensures that integration over $\theta$ is equivalent to integration over $a_0$.

---

## 6. Likelihood and Prior

Same as before:

### Likelihood:

$$
\log L(a_0) = \sum_{i=1}^n \left[ y_i a_0 - \log(1 + e^{a_0}) \right]
$$

### Prior:

$$
a_0 \sim \mathcal{N}(0, 10^2)
$$

### Combined:

$$
\log f(a_0) = \log L(a_0) + \log p(a_0)
$$
As giiven by

```python
logDens = logL + prior_a0
```

And

$$
\log f^*(\theta) = \log f(a_0(\theta)) + \log \left| \frac{da_0}{d\theta} \right|
$$
That is,

```python
logL = logDens+jacobianL
```

---

## 7. Clipping for Numerical Stability

The code enforces:

$$
\theta \in [-0.99999, 0.99999]
$$

This avoids:

* Division by zero at $ t = \pm 1 $
* Exploding Jacobian values

---

## 8. Integration Domain

Instead of:

$$
a_0 \in [-2, 1]
$$

we now integrate:

$$
t \in [-1, 1]
$$

This covers the **entire real line in ( a_0 )**.

---

## 9. Shifting Logic (Same idea as before)

Define:

$$
\ell_{\max} = \max_t \log f^*(t)
$$

Then integrate:

$$
\text{scaled integral} = \int \exp(\log f^*(t) - \ell_{\max}), dt
$$

Recovers log evidence as:

$$
\log I = \ell_{\max} + \log(\text{scaled integral})
$$

---

## 10. Posterior Computation

To compute the posterior and map it mack to the original scale, two functions i.e. `jac_trans` and `compute_log_lik_adj2` are defined. `compute_log_lik_adj2` defines the desnity without Jacobean, i.e., $\log f(a_0)=\log L(a_0)+\log p(a_0)$. With this and the log evidence $\log I$ obtained above, the normalized posterior can be easily obtained as 

$$
p(a_0∣y)= exp(f(a0) - \log I)
$$

Now the only step left is $t \mapsto a_0$ mapping, which the `jac_trans` function does using our original equation

$$
a_0 = \frac{\theta}{1 - \theta^2}, \quad t \in (-1,1)
$$




