---
title: "Vegas"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Two-arm trial TFP v RStan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

## Quick start
The model implemented here is a non-centered parameterization for a logistic model.


```{r rsetup, include=FALSE,eval=TRUE}
library(reticulate)
use_virtualenv("/Users/gogol/py_env/bayes", required = TRUE)
```

## Data set

The R chunk below creates a simple dataset of three cols:

-   a binary response variable (0/1 = non-responder/responder),
-   a binary treatment variable (0/1 = control/test treatment)
-   and a basket ID variable. (1)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. a classical two arm randomized trial design. Baskets will be added in other vignettes.

```{r datasets, include=TRUE,eval=TRUE}
### Prepare model inputs
set.seed(99999)
# Set up data
rr_k_ctrl <- c(0.20)        # control response rate for each basket
rr_k_trt <- c(0.40)         # treatment response rate for each basket

K<-length(rr_k_ctrl)        # number of baskets

N_k_ctrl <- rep(1500, K)     # number of control participants per basket
N_k_trt <- rep(1500, K)      # number of treatment participants per basket
N_k <- N_k_ctrl + N_k_trt   # number of participants per basket (both arms combined)
N <- sum(N_k)               # total sample size
k_vec <- rep(1:K, N_k)      # N x 1 vector of basket indicators (1 to K)

T_vec<-NULL;
y<-NULL;
for(i in 1:K){ # for each basket repeat 0-control 1-trt according to the specifc Ns
  T_vec<-c(T_vec,rep(0:1,c(N_k_ctrl[i],N_k_trt[i]))) # treatment/control indicator
  y<-c(y,
       c(rbinom(N_k_ctrl[i],1,rr_k_ctrl[i]), # bernoulli for control
         rbinom(N_k_trt[i],1,rr_k_trt[i]))) #           for trt
}

thedata<-data.frame(y,basketID=k_vec,Treatment=T_vec)

py$thedata<-r_to_py(thedata) # THE KEY LINE - makes data available to Python
                             # this is converted into a pandas dataframe object in Python

eta=0.1+0.5*T_vec[1:10]
p=exp(eta)/(1+exp(eta))

sum(y[1:10]*log(p)+(1-y[1:10])*log(1-p))

eta=-1.1+1.5*T_vec[1:10]
p=exp(eta)/(1+exp(eta))

exp(sum(y[1:10]*log(p)+(1-y[1:10])*log(1-p)))

summary(m1<-glm(y~1+Treatment,data=thedata[,],family="binomial"))

summary(m0<-glm(y~1,data=thedata[,],family="binomial"))

```

### Setup
```{python onevar,include=TRUE,eval=TRUE}
import numpy as np
import pandas as pd
#from jax import config
#config.update("jax_enable_x64", True)
import jax.numpy as jnp
import jax
import vegas
import math
import gvar

## The data.frame passed is now a panda df but for TPF this needs to be further
## converted into tensors here we simply convert each col of the data.frame into
## a Rank 1 tensor (i.e. a vector)
y_data=jnp.array(thedata.iloc[:,0],dtype=jnp.float32)[:,None] # 1 col matrix
k_vec=jnp.array(thedata.iloc[:,1])-1
T_vec=jnp.array(thedata.iloc[:,2],dtype=jnp.float32)[:,None] # 1 col matrix


pars=jnp.array([[0.1,-1.1,2.1]],dtype=jnp.float32)
#pars=jnp.stack([pars, pars,pars], axis=1)


@jax.jit
def compute_log_lik(theta, y,T):
    #print(f"shape1={x.shape}\n")
    """
    logL = sum_i^n ( y_i*(a0+a1*T_i) - log(1+ exp(a0+a1*T_i) )  )
    Expects x to have shape (batch_size, DIM).
    """
    #print(f"shape={theta.shape}\n")
    a0=theta[0,:]#,:] # (3,)
    eta=a0# (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - jnp.log(1+jnp.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)),axis=0) #
    # now add the priors
    # simple N(0,sd=10)
    prior_a0=jax.scipy.stats.norm.logpdf(a0,loc=0.,scale=10.)
    logDens = logL + prior_a0 
    return logDens # this will be (3,)

# 2. Setup Vegas with a Shifting Logic
@vegas.rbatchintegrand
class StableLikelihood:
    def __init__(self, y, T):
        self.y = y
        self.T = T
        self.l_max = 0.0 # Will be updated after warm-up

    def __call__(self, theta):
        theta_jax = jnp.array(theta)
        log_lik = compute_log_lik(theta_jax, self.y,self.T)
        
        # Shift by the maximum found so far
        # This keeps the result near 1.0 instead of 1e-300
        return jnp.exp(log_lik - self.l_max)

model = StableLikelihood(y=y_data,T=T_vec)

start = jnp.array([-2],dtype=jnp.float32) # a0 a1
stop = jnp.array([1],dtype=jnp.float32)   # a0 a1
searchPts=jnp.transpose(jnp.linspace(start, stop, 500))

model.l_max=jnp.max(compute_log_lik(searchPts, model.y,model.T))

#explicitly set seed
gvar.ranseed(9999)

integ = vegas.Integrator(jnp.array([[-2, 1]]*1,dtype=jnp.float32))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=10000)
# STAGE 2: Production (Stable Integration)
result = integ(model, nitn=10, neval=50000)

# 5. Output results
print(f"\nIntegration Results:")
print(f"Result: {result.mean:.8f} +/- {result.sdev:.8f}")
print(f"Q-value: {result.Q:.2f} (should be close to 0.5 for a good fit)")

# STAGE 3: Final Log-Evidence Calculation
# log(I) = l_max + log(scaled_integral)
log_evidence = model.l_max + jnp.log(result.mean)

print(f"Log Evidence: {log_evidence}")

# we want posterior values
# compute_log_lik(theta, y,T) - log_evidence = log( f(D|a0)f(a0) )

mydens=np.array(jnp.exp(compute_log_lik(searchPts,y_data,T_vec) - log_evidence))
myX=np.array(searchPts)

```
```{r onevarR,include=TRUE,eval=TRUE}

plot(py$myX,py$mydens,type="l")
abline(v=coef(m0))

```


```{python onevar_withjaco,include=TRUE,eval=TRUE}
import numpy as np
import pandas as pd
import jax.numpy as jnp
import jax
import vegas
import math
import gvar

y_data=jnp.array(thedata.iloc[:,0],dtype=jnp.float32)[:,None] # 1 col matrix
k_vec=jnp.array(thedata.iloc[:,1])-1
T_vec=jnp.array(thedata.iloc[:,2],dtype=jnp.float32)[:,None] # 1 col matrix


pars=jnp.array([[-0.1,0.2,0.3],[0.01,-0.9,0.05]],dtype=jnp.float32)
#pars=jnp.stack([pars, pars,pars], axis=1)

#change of variables, we take a0 and a1 both on -infty to + infty
# to a0 = t1/(1-t1^2) and a1 = t2/(1-t2^2)
# and jacobian is a = (1+t^2)/(1-t^2)^2
# so evaluate function at these transformations and add in jacobian terms


@jax.jit
def compute_log_lik_adj(theta, y,T):
    #print(f"shape1={x.shape}\n")
    """
    logL = sum_i^n ( y_i*(a0+a1*T_i) - log(1+ exp(a0+a1*T_i) )  )
    Expects x to have shape (batch_size, DIM).
    """
    #print(f"shape={theta.shape}\n")
    #a0=theta[0,:]#,:] # (3,)
    #a1=theta[1,:]#,:] # (3,)
    
    theta = jnp.clip(theta, -0.99999, 0.99999) 
    
    a0=theta[0,:]/(1-jnp.power(theta[0,:],2))
    
    jacobianL = jnp.log(1+jnp.power(theta[0,:],2)) - 2.0*jnp.log(1-jnp.power(theta[0,:],2))
    
    eta=a0 # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - jnp.log(1+jnp.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)),axis=0) #
    # now add the priors
    # simple N(0,sd=10)
    prior_a0=jax.scipy.stats.norm.logpdf(a0,loc=0.,scale=10.)
    logDens = logL + prior_a0
    logL = logDens+jacobianL
    return logL # this will be (3,)


# 2. Setup Vegas with a Shifting Logic
@vegas.rbatchintegrand
class StableLikelihood:
    def __init__(self, y, T):
        self.y = y
        self.T = T
        self.l_max = 0.0 # Will be updated after warm-up

    def __call__(self, theta):
        theta_jax = jnp.array(theta)
        log_lik = compute_log_lik_adj(theta_jax, self.y,self.T)
        
        # Shift by the maximum found so far
        # This keeps the result near 1.0 instead of 1e-300
        return jnp.exp(log_lik - self.l_max)

model = StableLikelihood(y=y_data,T=T_vec)

start = jnp.array([-1],dtype=jnp.float32) # a0 a1
stop = jnp.array([1],dtype=jnp.float32)   # a0 a1
searchPts=jnp.transpose(jnp.linspace(start, stop, 500,endpoint=False))

model.l_max=jnp.max(compute_log_lik_adj(searchPts, model.y,model.T))
#explicitly set seed
gvar.ranseed(9999)

integ = vegas.Integrator(jnp.array([[-1, 1]]*1,dtype=jnp.float32))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=1000)
# STAGE 2: Production (Stable Integration)
result = integ(model, nitn=10, neval=5000)

# 5. Output results
print(f"\nIntegration Results:")
print(f"Result: {result.mean:.8f} +/- {result.sdev:.8f}")
print(f"Q-value: {result.Q:.2f} (should be close to 0.5 for a good fit)")

# STAGE 3: Final Log-Evidence Calculation
# log(I) = l_max + log(scaled_integral)
log_evidence = model.l_max + jnp.log(result.mean)

print(f"Log Evidence: {log_evidence}")


start2 = jnp.array([-0.7],dtype=jnp.float32) # a0 a1
stop2 = jnp.array([0.2],dtype=jnp.float32)   # a0 a1
searchPts2=jnp.transpose(jnp.linspace(start2, stop2, 500,endpoint=False))


# to map from -1 +1 mapping to -inf + inf - searchPts2 on -1 +1
def jac_trans(theta):
    theta = jnp.clip(theta, -0.99999, 0.99999) 
    a0=theta[0,:]/(1-jnp.power(theta[0,:],2))
    return(a0)

# need to turn off jacobian in density - just need density after mapped from -1 +1 to -inf +inf
def compute_log_lik_adj2(theta, y,T):
    
    theta = jnp.clip(theta, -0.99999, 0.99999) 
    
    a0=theta[0,:]/(1-jnp.power(theta[0,:],2))
    
    eta=a0 # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - jnp.log(1+jnp.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)),axis=0) #
    # now add the priors
    # simple N(0,sd=10)
    prior_a0=jax.scipy.stats.norm.logpdf(a0,loc=0.,scale=10.)
    logDens = logL + prior_a0
    logL = logDens
    return logL # this will be (3,)

    
# we want posterior values
# compute_log_lik(theta, y,T) - log_evidence = log( f(D|a0,a1)f(a0)f(a1) )
mydens2=np.array(jnp.exp(compute_log_lik_adj2(searchPts2,y_data,T_vec) - log_evidence))
myX2=np.array(jac_trans(searchPts2))


```

```{r onevarR2,include=TRUE,eval=TRUE}

par(mfrow=c(1,2))
plot(py$myX,py$mydens,type="l")
abline(v=coef(m0))
plot(py$myX2,py$mydens2,type="l",col="red")
abline(v=coef(m0))

```


```{python twovar,include=TRUE,eval=TRUE}
import numpy as np
import pandas as pd
from jax import config
config.update("jax_enable_x64", True)
import jax.numpy as jnp
import jax
import vegas
import math
import gvar

## The data.frame passed is now a panda df but for TPF this needs to be further
## converted into tensors here we simply convert each col of the data.frame into
## a Rank 1 tensor (i.e. a vector)
y_data=jnp.array(thedata.iloc[:,0],dtype=jnp.float64)[:,None] # 1 col matrix
k_vec=jnp.array(thedata.iloc[:,1])-1
T_vec=jnp.array(thedata.iloc[:,2],dtype=jnp.float64)[:,None] # 1 col matrix


pars=jnp.array([[0.1,-1.1,2.1],[0.5,1.5,-2.5]],dtype=jnp.float64)
#pars=jnp.stack([pars, pars,pars], axis=1)


@jax.jit
def compute_log_lik(theta, y,T):
    #print(f"shape1={x.shape}\n")
    """
    logL = sum_i^n ( y_i*(a0+a1*T_i) - log(1+ exp(a0+a1*T_i) )  )
    Expects x to have shape (batch_size, DIM).
    """
    #print(f"shape={theta.shape}\n")
    a0=theta[0,:]#,:] # (3,)
    a1=theta[1,:]#,:] # (3,)
    eta=a0+a1*T # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - jnp.log(1+jnp.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)),axis=0) #
    # now add the priors
    # simple N(0,sd=10)
    prior_a0=jax.scipy.stats.norm.logpdf(a0,loc=0.,scale=10.)
    prior_a1=jax.scipy.stats.norm.logpdf(a1,loc=0.,scale=10.)
    logDens = logL + prior_a0 + prior_a1
    return logDens # this will be (3,)

# 2. Setup Vegas with a Shifting Logic
@vegas.rbatchintegrand
class StableLikelihood:
    def __init__(self, y, T):
        self.y = y
        self.T = T
        self.l_max = 0.0 # Will be updated after warm-up

    def __call__(self, theta):
        theta_jax = jnp.array(theta)
        log_lik = compute_log_lik(theta_jax, self.y,self.T)
        
        # Shift by the maximum found so far
        # This keeps the result near 1.0 instead of 1e-300
        return np.array(jnp.exp(log_lik - self.l_max)).copy()

model = StableLikelihood(y=y_data,T=T_vec)

start = jnp.array([-10, -10],dtype=jnp.float64) # a0 a1
stop = jnp.array([10, 10],dtype=jnp.float64)   # a0 a1
searchPts=jnp.transpose(jnp.linspace(start, stop, 500))

model.l_max=jnp.max(compute_log_lik(searchPts, model.y,model.T))

#explicitly set seed
gvar.ranseed(9999)

integ = vegas.Integrator(jnp.array([[-10, 10]]*2,dtype=jnp.float64))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=10000)
# STAGE 2: Production (Stable Integration)
result = integ(model, nitn=10, neval=50000)

# 5. Output results
print(f"\nIntegration Results:")
print(f"Result: {result.mean:.8f} +/- {result.sdev:.8f}")
print(f"Q-value: {result.Q:.2f} (should be close to 0.5 for a good fit)")

# STAGE 3: Final Log-Evidence Calculation
# log(I) = l_max + log(scaled_integral)
log_evidence = model.l_max + jnp.log(result.mean)

print(f"Log Evidence: {log_evidence}")


```

```{python twovar_withjaco,include=TRUE,eval=TRUE}
import numpy as np
import pandas as pd
import jax.numpy as jnp
import jax
import vegas
import math
import gvar
from jax import config
config.update("jax_enable_x64", True)

y_data=jnp.array(thedata.iloc[:,0],dtype=jnp.float64)[:,None] # 1 col matrix
k_vec=jnp.array(thedata.iloc[:,1])-1
T_vec=jnp.array(thedata.iloc[:,2],dtype=jnp.float64)[:,None] # 1 col matrix


pars=jnp.array([[-0.1,0.2,0.3],[0.01,-0.9,0.05]],dtype=jnp.float64)
#pars=jnp.stack([pars, pars,pars], axis=1)

#change of variables, we take a0 and a1 both on -infty to + infty
# to a0 = t1/(1-t1^2) and a1 = t2/(1-t2^2)
# and jacobian is a = (1+t^2)/(1-t^2)^2
# so evaluate function at these transformations and add in jacobian terms


@jax.jit
def compute_log_lik_adj(theta, y,T):
    #print(f"shape1={x.shape}\n")
    """
    logL = sum_i^n ( y_i*(a0+a1*T_i) - log(1+ exp(a0+a1*T_i) )  )
    Expects x to have shape (batch_size, DIM).
    """
    #print(f"shape={theta.shape}\n")
    #a0=theta[0,:]#,:] # (3,)
    #a1=theta[1,:]#,:] # (3,)
    
    theta = jnp.clip(theta, -0.99999, 0.99999) 
    
    a0=theta[0,:]/(1-jnp.power(theta[0,:],2))
    a1=theta[1,:]/(1-jnp.power(theta[1,:],2))
    
    jacobianL = jnp.log(1+jnp.power(theta[0,:],2)) - 2.0*jnp.log(1-jnp.power(theta[0,:],2))+jnp.log(1+jnp.power(theta[1,:],2)) - 2.0*jnp.log(1-jnp.power(theta[1,:],2))
    
     
    eta=a0+a1*T # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - jnp.log(1+jnp.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = jnp.sum(y*eta - jnp.log(1+jnp.exp(eta)),axis=0) #
    # now add the priors
    # simple N(0,sd=10)
    prior_a0=jax.scipy.stats.norm.logpdf(a0,loc=0.,scale=10.)
    prior_a1=jax.scipy.stats.norm.logpdf(a1,loc=0.,scale=10.)
    logDens = logL + prior_a0 + prior_a1
    logLL = logDens+jacobianL
    #print(f"theta={theta}\n\na0={a0} a1={a1} logL={logL} expeta={jnp.exp(eta)} jacoL={jacobianL}\n")
    return logL # this will be (3,)


# 2. Setup Vegas with a Shifting Logic
@vegas.rbatchintegrand
class StableLikelihood:
    def __init__(self, y, T):
        self.y = y
        self.T = T
        self.l_max = 0.0 # Will be updated after warm-up

    def __call__(self, theta):
        theta_jax = jnp.array(theta)
        log_lik = compute_log_lik_adj(theta_jax, self.y,self.T)
        
        # Shift by the maximum found so far
        # This keeps the result near 1.0 instead of 1e-300
        #print(f"A=={log_lik} B= {self.l_max}\n")
        return np.array(jnp.exp(log_lik - self.l_max)).copy()

model = StableLikelihood(y=y_data,T=T_vec)

start = jnp.array([-1, -1],dtype=jnp.float64) # a0 a1
stop = jnp.array([1, 1],dtype=jnp.float64)   # a0 a1
searchPts=jnp.transpose(jnp.linspace(start, stop, 5000,endpoint=False))

model.l_max=jnp.max(compute_log_lik_adj(searchPts, model.y,model.T))
#explicitly set seed
gvar.ranseed(9999)

integ = vegas.Integrator(jnp.array([[-1, 1]]*2,dtype=jnp.float64))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=10000)
# STAGE 2: Production (Stable Integration)
result = integ(model, nitn=10, neval=50000)


# 5. Output results
print(f"\nIntegration Results:")
print(f"Result: {result.mean:.8f} +/- {result.sdev:.8f}")
print(f"Q-value: {result.Q:.2f} (should be close to 0.5 for a good fit)")

# STAGE 3: Final Log-Evidence Calculation
# log(I) = l_max + log(scaled_integral)
log_evidence = model.l_max + jnp.log(result.mean)

print(f"Log Evidence: {log_evidence}")


# we want posterior values
# compute_log_lik(theta, y,T) - log_evidence = log( f(D|a0,a1)f(a0)f(a1) )


```



