---
title: "Logistic BHM - Vegas MLX v RStan"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Two-arm trial TFP v RStan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
#on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

# If on GitHub, set eval = FALSE for all subsequent chunks
#knitr::opts_chunk$set(eval = !on_github)

```

### Quickstart
This vignettes compares the same Bayesian hierarchical model estimated with Vegas and Rstan. This is a logistic regression model. Only key snippets and outputs are shown below. The full vignette is available [here](https://raw.githubusercontent.com/GSK-Biostatistics/vignettes-asia/refs/heads/main/vegas/stanvegas1_mlx.qmd)

This vignette is almost identical to the other vignette ``Logistic BHM - Vegas JAX v RStan'' but this one uses the MLX library for the calculations which uses a GPU, and is considerably faster the JAX equivalent but requires MacOS and Apple Silicon platform. 

### Model Formulation

The model implemented here is a non-centered parameterization (helps improve MCMC sampling) for a logistic model.

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\tilde{\theta}_j &\sim \text{Normal}(0, 1)\quad\enspace\text{for}\enspace{j=1,2}  \\
\beta_0 &= \mu_0 + \sigma_0 \tilde{\theta}_1 \\
\beta_1 &= \mu_1 + \sigma_1 \tilde{\theta}_2 \\
\text{logit(}{p_i}) &= \beta_0 + \beta_1 z_i\quad\quad\enspace\enspace\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$
When using direct integration via Vegas we don't need the non-centered parameterization:

$$
\begin{aligned}
\beta_0 &\sim \text{Normal}(\mu_0, \sigma_0) \\
\beta_1 &\sim \text{Normal}(\mu_1, \sigma_1) \\
\end{aligned}
$$

```{r rsetup, include=FALSE,eval=TRUE}
if(Sys.info()["sysname"]=="Darwin" && Sys.info()["machine"]=="arm64"){
  cat("MacOS and Apple ARM detected\n")
  } else {cat("WARNING: This vignette is for MLX (Apple ARM Platform)")}
library(reticulate)
# use built-in reticulate environment - can use custom if desired - see use_virtualenv
install_if_missing <- function(packages) {
  for (pkg in packages) {
    if (!reticulate::py_module_available(pkg)) {
      reticulate::py_install(pkg)
    }
  }
}
install_if_missing(c("numpy","pandas","jax","vegas","gvar","mlx","pickle"))
#use_virtualenv("/mnt/myvenv") # or use custom venv
library(rstan)
library(bayesplot)
library(ggplot2)
library(kableExtra)
#reticulate::py_install("bash") # just a test
```

## Data set

The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a (dummy) basket ID variable (=1)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. a classical two arm randomized trial design. Baskets will be added in other vignettes.

```{r datasets, include=FALSE,eval=TRUE}
### Prepare model inputs
set.seed(99999)
# Set up data
rr_k_ctrl <- c(0.20)        # control response rate for each basket
rr_k_trt <- c(0.40)         # treatment response rate for each basket

K<-length(rr_k_ctrl)        # number of baskets

N_k_ctrl <- rep(100, K)     # number of control participants per basket
N_k_trt <- rep(100, K)      # number of treatment participants per basket
N_k <- N_k_ctrl + N_k_trt   # number of participants per basket (both arms combined)
N <- sum(N_k)               # total sample size
k_vec <- rep(1:K, N_k)      # N x 1 vector of basket indicators (1 to K)

z_vec<-NULL;
y<-NULL;
for(i in 1:K){ # for each basket repeat 0-control 1-trt according to the specifc Ns
  z_vec<-c(z_vec,rep(0:1,c(N_k_ctrl[i],N_k_trt[i]))) # treatment/control indicator
  y<-c(y,
       c(rbinom(N_k_ctrl[i],1,rr_k_ctrl[i]), # bernoulli for control
         rbinom(N_k_trt[i],1,rr_k_trt[i]))) #           for trt
}

thedata<-data.frame(y,basketID=k_vec,Treatment=z_vec)

py$thedata<-r_to_py(thedata) # THE KEY LINE - makes data available to Python
                             # this is converted into a pandas dataframe object in Python
```



## Vegas

```{python evid_and_a0,include=FALSE,eval=TRUE}
import numpy as np
import pandas as pd
#from jax import config
#config.update("jax_enable_x64", True)  ## if turn this on then vegas need np.array() in function return values
import jax.numpy as jnp
import jax
import vegas
import math
import gvar
import mlx 
import mlx.core as mx
import pickle

## The data.frame passed is now a panda df needs converted to mlx
y_data=mx.array(thedata.iloc[:, 0].to_numpy(), dtype=mx.float32)[:,None]
k_vec=mx.array(thedata.iloc[:, 1].to_numpy())-1
T_vec=mx.array(thedata.iloc[:, 2].to_numpy(), dtype=mx.float32)[:,None]

#with open("y_data.pkl","wb") as file:
#  pickle.dump(y_data,file)
#with open("k_vec.pkl","wb") as file:
#  pickle.dump(k_vec,file)
#with open("T_vec.pkl","wb") as file:
#  pickle.dump(T_vec,file)

#### ------------------------------------------------------------------------------------------------
#### 1. DEFINE HELPER FUNCTIONS - STATS FUNCTIONS ARE LATER
def converged(vegasresult, reltol, Qtol):
    res = vegasresult # note this is results object OR vegas.ravg object - both have necessary members
    # This returns True if the condition is met, False otherwise
    return (100 * (gvar.sdev(res) / gvar.mean(res)) < reltol) and (res.Q > Qtol)

def runIters(integ,model,warm_nitn,warm_neval,run_nitn,run_neval,iMax,errTol,Qtol,verbose=False):
    # RUN VEGAS to hit a given accuracy level AND checks Q
    # this runs warmup and sets of iters until specified error is met, 100*sdev/mean < errTol
    # e.g. errTol = 0.1 is sd is less than 0.1% of mean
    # returns a list comprising of integ() object and a boolean, boolean is to check accuracy was hit
    # Warmup grid - do not store results with rest as early grid selection
    res_warmup = integ(model, nitn=warm_nitn, neval=warm_neval)
    # Production (Stable Integration)
    results1 = integ(model, nitn=run_nitn, neval=run_neval)
    # now check if the desired accuracy is achieved and Q is not too low
    i=0
    results1itn=results1.itn_results # pull out just iters as array
    success = True
    while (not converged(results1,errTol,Qtol)) and i<iMax:
        if verbose:
            print(f"cumulative iterations={i*run_nitn+run_nitn}\n")
        # not converged yet so run another set of iterations - each updates grid
        results2 = integ(model, nitn=run_nitn, neval=run_neval)
        # combine the results with existing set
        results1itn = results1itn+results2.itn_results #existing plus newest
        results1=vegas.ravg(results1itn) # this is then passed to while condition to see if more integ() needed
        i=i+1 # counter to control too many iters
        if i==iMax:
            success=False
    return list([results1,success])

```

To use direct integration in Vegas the key parts are:

-  define the prior density functions
-  define the log likelihood + log priors

In addition, we also implement a **change of variables** to deal with the range of integration going to $\infty$. This means we integrate over transformed variables (see below) which have finite bounds. As we are changing variables then this requires the usual Jacobian adjustment to the function being integrated to ensure the volume being computed is equivalent to that on the original scale.

To map a variable $t\in(-1,1)$ to $x\in(-\infty,\infty)$ we can use the following function and it's Jacobian (J):
$$
\begin{aligned}
x&=&\frac{t}{1-t^2} \\
\\
\enspace\text{where } \quad\enspace J &=& \frac{1+t^2}{(1-t^2)^2}
\end{aligned}
$$

The function below `compute_log_lik(theta, y,T)` is necessary to compute the overall standardization constant - the constant which ensures the posterior density integrates to unity - and the integration here is over all parameters in the model. We work in log scale as much as possible and a scaling ``trick'' is used to help avoid numerical overflow. These technical numerical details can be found in the full quarto file.  

```{python sixvarjaco_a,include=TRUE,eval=TRUE}
#### ------------------------------------------------------------------------------------------------
#### 2. DEFINE STATS FUNCTIONS 
#### ------------------------------------------------------------------------------------------------

@mx.compile
def mx_half_normal_logpdf(x, sigma):
    # Log-constant: log(sqrt(2) / sqrt(pi))
    log_const = 0.5*(mx.log(2.0) - mx.log(mx.pi)) - mx.log(sigma)
    # Log-exponent term
    log_exp = -((x*x) / (2.0 * sigma*sigma))
    return log_const+log_exp

@mx.compile
def mx_norm_logpdf(x, loc, scale):
    # -0.5 * ((x - loc) / scale)**2 - log(scale) - 0.5 * log(2 * pi)
    log_var = -0.5 * ((x - loc) / scale)**2
    log_const = -mx.log(scale) - 0.5 * math.log(2 * math.pi)
    return log_var + log_const

### SETUP LIKELIHOOD*PRIOR 
@mx.compile
def compute_log_lik(theta, y,T):
    #print(f"shape1={x.shape}\n")
    """
    mu0 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
    sigma0 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
    mu1 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
    sigma1 ~ normal(0, 2.5);
    a0 ~ normal(mu0,sigma0)
    a1 ~ normal(mu1,sigma1)
    
    logL + logpdf(a0|mu0,sigma0) + logpdf(a1|mu1,sigma1) + logpdf(mu0) + logpdf(m1) + logpdf(sigma0) + logpdf(sigma1)
    
    logL = sum_i^n ( y_i*(a0+a1*T_i) - log(1+ exp(a0+a1*T_i) )  )
    logpdf(a0|mu0,sigma0) = jax.scipy.stats.norm.logpdf(a0,loc=mu0,scale=sigma0)
    logpdf(a1|mu1,sigma1) = jax.scipy.stats.norm.logpdf(a1,loc=mu1,scale=sigma1)
    logpdf(mu0) = jax.scipy.stats.norm.logpdf(mu0,loc=0.,scale=2.5)
    logpdf(mu1) = jax.scipy.stats.norm.logpdf(mu1,loc=0.,scale=2.5)
    logpdf(sigma0) = half_normal_logpdf(sigma0, 2.5)
    logpdf(sigma1) = half_normal_logpdf(sigma1, 2.5)
    Expects x to have shape (batch_size, DIM).
    
    """
    #print(f"shape={theta.shape}\n")
    #print(f"shape={theta.shape}\n")
    
    theta0=mx.clip(theta[0,:],-0.99999, 0.99999)
    theta1=mx.clip(theta[1,:],-0.99999, 0.99999)
    theta2=mx.clip(theta[2,:],-0.99999, 0.99999)
    theta3=mx.clip(theta[3,:],-0.99999, 0.99999)
    theta4=mx.clip(theta[4,:],0.00001, 0.99999)
    theta5=mx.clip(theta[5,:],0.00001, 0.99999)
    
    
    jacobianL = (  mx.log1p(mx.power(theta0,2)) - 2.0*mx.log1p(-mx.power(theta0,2)) 
                 + mx.log1p(mx.power(theta1,2)) - 2.0*mx.log1p(-mx.power(theta1,2))
                 + mx.log1p(mx.power(theta2,2)) - 2.0*mx.log1p(-mx.power(theta2,2))
                 + mx.log1p(mx.power(theta3,2)) - 2.0*mx.log1p(-mx.power(theta3,2))
                 + mx.log1p(mx.power(theta4,2)) - 2.0*mx.log1p(-mx.power(theta4,2))
                 + mx.log1p(mx.power(theta5,2)) - 2.0*mx.log1p(-mx.power(theta5,2))
                 )
    #print(f"jacobian={jacobianL}\n")             

    a0=theta0/(1-mx.power(theta0,2))
    a1=theta1/(1-mx.power(theta1,2))
    mu0=theta2/(1-mx.power(theta2,2))
    mu1=theta3/(1-mx.power(theta3,2))
    sigma0=theta4/(1-mx.power(theta4,2))
    sigma1=theta5/(1-mx.power(theta5,2))
    #print(f"a0={a0} a1={a1} mu0={mu0} mu1={mu1} sigma0={sigma0} sigma1={sigma1}\n")
    
    eta=a0+a1*T # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - mx.log(1+mx.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = mx.sum(y*eta - mx.log1p(mx.exp(eta)),axis=0) #
    #print(f"logL={logL}\n")
    # now add the priors
    # simple N(0,sd=10)
    prior_a0 = mx_norm_logpdf(a0,loc=mu0,scale=sigma0)
    prior_a1 = mx_norm_logpdf(a1,loc=mu1,scale=sigma1)
    prior_mu0 = mx_norm_logpdf(mu0,loc=0.,scale=2.5)
    prior_mu1 = mx_norm_logpdf(mu1,loc=0.,scale=2.5)
    prior_sigma0 = mx_half_normal_logpdf(sigma0, 2.5)
    prior_sigma1 = mx_half_normal_logpdf(sigma1, 2.5)
    
    logDens = logL + prior_a0 + prior_a1 + prior_mu0 + prior_mu1 + prior_sigma0 + prior_sigma1
    #print(f"logDens={logDens}\n")
    return (logDens + jacobianL) # this will be (3,)
  
```



```{python sixvarjaco_b,include=FALSE,eval=TRUE}
# 2. Setup Vegas with a Shifting Logic
@vegas.rbatchintegrand
class StableLikelihood:
    def __init__(self, y, T):
        self.y = y
        self.T = T
        self.l_max = 0.0 # Will be updated after warm-up

    def __call__(self, theta):
        theta_mx = mx.array(theta)
        log_lik = compute_log_lik(theta_mx, self.y,self.T)
        
        # Shift by the maximum found so far
        # This keeps the result near 1.0 instead of 1e-300
        ## return np.array(jnp.exp(log_lik - self.l_max)) # needt this is using x64 - vegas expects np
        return mx.exp(log_lik - self.l_max)

model = StableLikelihood(y=y_data,T=T_vec)

start = jnp.array([-1,-1,-1,-1,0.0001,0.0001],dtype=jnp.float32) # a0 a1
stop = jnp.array([1,1,1,1,1,1],dtype=jnp.float32)   # a0 a1
searchPts=mx.array(np.array(jnp.transpose(jnp.linspace(start, stop, 1000))), dtype=mx.float32)

model.l_max=mx.max(compute_log_lik(searchPts, model.y,model.T))

#explicitly set seed
gvar.ranseed(99990)

integ = vegas.Integrator(mx.array([[-1, 1],[-1,1],[-1,1],[-1,1],[0.0001,1],[0.0001,1]],dtype=mx.float32))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=10000)
# STAGE 2: Production (Stable Integration)
result = integ(model, nitn=10, neval=50000)

# 5. Output results
#print(f"\nIntegration Results:")
#print(result.summary())
#print(f"Result: {result.mean:.8f} +/- {result.sdev:.8f}")
#print(f"Q-value: {result.Q:.2f} (should be close to 0.5 for a good fit)")

# STAGE 3: Final Log-Evidence Calculation
# log(I) = l_max + log(scaled_integral)
log_evidence = model.l_max + mx.log(result.mean)

print(f"Log Evidence: {log_evidence}")

# we want posterior values
# compute_log_lik(theta, y,T) - log_evidence = log( f(D|a0)f(a0) )

#mydens=np.array(jnp.exp(compute_log_lik(searchPts,y_data,T_vec) - log_evidence))
#myX=np.array(searchPts)

```

```{python genz,include=FALSE,eval=TRUE}
import mlx.core as mx
import numpy as np
import heapq
import itertools
import time
import mlx 
#import pickle

#with open("y_data.pkl","rb") as file:
#  loaded_data=pickle.load(file)
#with open("k_vec.pkl","rb") as file:
#  loaded_data=pickle.load(file)
#with open("T_vec.pkl","rb") as file:
#  loaded_data=pickle.load(file)

T=T_vec
y=y_data
mymax=model.l_max

# 1. GPU-Optimized Integrand
@mx.compile
def integrand_gpu(pts):
    theta=mx.transpose(pts)
    """
    Evaluates a 6D Gaussian peak centered at (0.5, ..., 0.5).
    'pts' is a MLX array of shape (N, 6).
    """
    #print(f"shape of pts={pts.shape}\n")
    theta0=mx.clip(theta[0,:],-0.99999, 0.99999)
    theta1=mx.clip(theta[1,:],-0.99999, 0.99999)
    theta2=mx.clip(theta[2,:],-0.99999, 0.99999)
    theta3=mx.clip(theta[3,:],-0.99999, 0.99999)
    theta4=mx.clip(theta[4,:],0.00001, 0.99999)
    theta5=mx.clip(theta[5,:],0.00001, 0.99999)
    
    
    jacobianL = (  mx.log1p(mx.power(theta0,2)) - 2.0*mx.log1p(-mx.power(theta0,2)) 
                 + mx.log1p(mx.power(theta1,2)) - 2.0*mx.log1p(-mx.power(theta1,2))
                 + mx.log1p(mx.power(theta2,2)) - 2.0*mx.log1p(-mx.power(theta2,2))
                 + mx.log1p(mx.power(theta3,2)) - 2.0*mx.log1p(-mx.power(theta3,2))
                 + mx.log1p(mx.power(theta4,2)) - 2.0*mx.log1p(-mx.power(theta4,2))
                 + mx.log1p(mx.power(theta5,2)) - 2.0*mx.log1p(-mx.power(theta5,2))
                 )
    #print(f"jacobian={jacobianL}\n")             

    a0=theta0/(1-mx.power(theta0,2))
    a1=theta1/(1-mx.power(theta1,2))
    mu0=theta2/(1-mx.power(theta2,2))
    mu1=theta3/(1-mx.power(theta3,2))
    sigma0=theta4/(1-mx.power(theta4,2))
    sigma1=theta5/(1-mx.power(theta5,2))
    #print(f"a0={a0} a1={a1} mu0={mu0} mu1={mu1} sigma0={sigma0} sigma1={sigma1}\n")
    
    eta=a0+a1*T # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - mx.log(1+mx.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = mx.sum(y*eta - mx.log1p(mx.exp(eta)),axis=0) #
    #print(f"logL={logL}\n")
    # now add the priors
    # simple N(0,sd=10)
    prior_a0 = mx_norm_logpdf(a0,loc=mu0,scale=sigma0)
    prior_a1 = mx_norm_logpdf(a1,loc=mu1,scale=sigma1)
    prior_mu0 = mx_norm_logpdf(mu0,loc=0.,scale=2.5)
    prior_mu1 = mx_norm_logpdf(mu1,loc=0.,scale=2.5)
    prior_sigma0 = mx_half_normal_logpdf(sigma0, 2.5)
    prior_sigma1 = mx_half_normal_logpdf(sigma1, 2.5)
    
    logDens = logL + prior_a0 + prior_a1 + prior_mu0 + prior_mu1 + prior_sigma0 + prior_sigma1
    #print(f"logDens={logDens}\n")
    print(f"mymax={mymax}\n")
    return mx.exp(logDens + jacobianL - mymax) # this will be (3,)

#start = jnp.array([-1,-1,-1,-1,0.0001,0.0001],dtype=jnp.float32) # a0 a1
#stop = jnp.array([1,1,1,1,1,1],dtype=jnp.float32)   # a0 a1
#searchPts=mx.array(np.array(jnp.transpose(jnp.linspace(start, stop, 1000))), dtype=mx.float32)

#mx.max(integrand_gpu(mx.transpose(searchPts)))


# 2. Genz-Malik Degree 7 Rule for n=6
n = 6
lam2 = np.sqrt(9/70)
lam4 = np.sqrt(9/10)
lam5 = np.sqrt(9/19)

nodes = []
nodes.append([0.0] * n) # Center (1 point)

for i in range(n): # Axial lam2 (2n points)
    p1, p2 = [0.0] * n, [0.0] * n
    p1[i], p2[i] = lam2, -lam2
    nodes.append(p1); nodes.append(p2)

for i in range(n): # Axial lam4 (2n points)
    p1, p2 = [0.0] * n, [0.0] * n
    p1[i], p2[i] = lam4, -lam4
    nodes.append(p1); nodes.append(p2)

for i, j in itertools.combinations(range(n), 2): # Cross lam4 (2n(n-1) points)
    for si, sj in [(1,1), (1,-1), (-1,1), (-1,-1)]:
        p = [0.0] * n
        p[i], p[j] = si * lam4, sj * lam4
        nodes.append(p)

for signs in itertools.product([1.0, -1.0], repeat=n): # Diagonal lam5 (2^n points)
    nodes.append([s * lam5 for s in signs])

REL_PTS = mx.array(nodes) 
num_points = REL_PTS.shape[0]
print(f"Generated {num_points} points for {n}D Genz-Malik rule.")

# Pre-calculate Weights as MLX arrays
w1 = (12824 - 9120*n + 400*n**2) / 19683
w2, w4 = 980 / 6561, 200 / 19683
w3, w5 = (1820 - 400*n) / 19683, 6859 / (19683 * 2**n)
WEIGHTS_7 = mx.array([w1] + [w2]*(2*n) + [w3]*(2*n) + [w4]*(2*n*(n-1)) + [w5]*(2**n))

we1 = (729 - 950*n + 50*n**2) / 729
we2, we4 = 245 / 486, 25 / 729
we3, we5 = (265 - 100*n) / 1458, 0.0
WEIGHTS_5 = mx.array([we1] + [we2]*(2*n) + [we3]*(2*n) + [we4]*(2*n*(n-1)) + [we5]*(2**n))

@mx.compile
def evaluate_batch_gpu(m, h):
    """
    Fully vectorized evaluation of B regions on GPU.
    """
    # Broadcasting: (B, 1, n) + (1, N, n) * (B, 1, n) -> (B, N, n)
    pts = m[:, None, :] + REL_PTS[None, :, :] * h[:, None, :]
    
    # Flatten and evaluate
    vals = integrand_gpu(mx.reshape(pts, (-1, n)))
    vals = mx.reshape(vals, (-1, num_points))
    
    # Calculate volumes (B,)
    volumes = mx.prod(2.0 * h, axis=1)
    
    # Calculate Degree 7 and Degree 5 integrals
    i7 = mx.sum(vals * WEIGHTS_7, axis=1) * volumes
    i5 = mx.sum(vals * WEIGHTS_5, axis=1) * volumes
    errs = mx.abs(i7 - i5)
    
    return i7, errs

def adaptive_integration_with_grid(initial_grid=None, bounds=None, tol=1e-3, batch_size=1024*4):
    """
    initial_grid: NumPy array of shape (num_regions, 2, n_dim)
                  grid[i, 0, :] are lower bounds of region i
                  grid[i, 1, :] are upper bounds of region i
                  
    low, high = np.array(bounds[0]), np.array(bounds[1])
    h_init = (high - low) / 2
    m_init = (low + high) / 2
    
    # Initial Evaluation
    m_mx = mx.array([m_init.tolist()])
    h_mx = mx.array([h_init.tolist()])
    i7, errs = evaluate_batch_gpu(m_mx, h_mx)
    
    total_integral = i7[0].item()
    total_error = errs[0].item()
    
    # Queue stores: (-error, center_list, halfwidth_list, integral)
    queue = [(-total_error, m_init.tolist(), h_init.tolist(), total_integral)]
                  
                  
                  
    """
    if initial_grid is not None:
        lows = initial_grid[:, 0, :]
        highs = initial_grid[:, 1, :]
    else:
        lows = np.array([bounds[0]])
        highs = np.array([bounds[1]])

    m_init_np = (lows + highs) / 2
    h_init_np = (highs - lows) / 2

    print(f"Initializing with {len(m_init_np)} regions...")
    
    # 1. Initial GPU Evaluation
    m_mx, h_mx = mx.array(m_init_np.tolist()), mx.array(h_init_np.tolist())
    i7_batch, err_batch = evaluate_batch_gpu(m_mx, h_mx)
    mx.eval(i7_batch, err_batch)
    
    i7_np, err_np = np.array(i7_batch), np.array(err_batch)
    total_integral = np.sum(i7_np)
    total_error = np.sum(err_np)
    
    queue = []
    for i in range(len(m_init_np)):
        heapq.heappush(queue, (-err_np[i], m_init_np[i].tolist(), h_init_np[i].tolist(), i7_np[i]))

    start_time = time.time()
    iterations = 0
    
    print(f"Starting {n}D Adaptive Integration (GPU Optimized)")
    
    while total_error > tol and len(queue) > 0:
        iterations += 1
        
        # 1. Select regions from priority queue
        curr_batch = min(len(queue), batch_size)
        m_list, h_list, p_ints, p_errs = [], [], 0.0, 0.0
        
        for _ in range(curr_batch):
            e_neg, m_c, h_c, i_c = heapq.heappop(queue)
            m_list.append(m_c)
            h_list.append(h_c)
            p_ints += i_c
            p_errs += (-e_neg)
            
        # 2. Split regions (CPU logic)
        m_arr = np.array(m_list)
        h_arr = np.array(h_list)
        split_dims = np.argmax(h_arr, axis=1)
        
        m_children, h_children = [], []
        for i in range(curr_batch):
            d = split_dims[i]
            h_new = h_arr[i].copy()
            h_new[d] /= 2
            
            m1, m2 = m_arr[i].copy(), m_arr[i].copy()
            m1[d] -= h_new[d]
            m2[d] += h_new[d]
            
            m_children.append(m1.tolist()); m_children.append(m2.tolist())
            h_children.append(h_new.tolist()); h_children.append(h_new.tolist())
            
        # 3. Evaluate Children on GPU
        m_mx = mx.array(m_children)
        h_mx = mx.array(h_children)
        i7_gpu, err_gpu = evaluate_batch_gpu(m_mx, h_mx)
        
        mx.eval(i7_gpu, err_gpu)
        i7_cpu = np.array(i7_gpu)
        err_cpu = np.array(err_gpu)
        
        # 4. Update state
        total_integral -= p_ints
        total_error -= p_errs
        
        for i in range(len(m_children)):
            integral, error = i7_cpu[i], err_cpu[i]
            total_integral += integral
            total_error += error
            heapq.heappush(queue, (-error, m_children[i], h_children[i], integral))
            
        if iterations % 100 == 0:
          # Peek at the top of the queue for the max regional error
            max_regional_error = -queue[0][0]
            print(f"Iteration {iterations:5d} | Regions: {len(queue):8d} | Total Err: {total_error:.4e} | Max Reg Err: {max_regional_error:.4e} | Integral: {total_integral:.6f}")
            #print(f"Iteration {iterations:5d} | Error: {total_error:.4e} | Integral: {total_integral:.8e}")
            
    print(f"\n--- {n}D Integration Summary ---")
    print(f"Total Iterations: {iterations}")
    print(f"Total Regions:    {len(queue)}")
    print(f"Final Integral:   {total_integral:.12e}")
    print(f"Final Error:      {total_error:.4e}")
    print(f"Time Taken:       {time.time() - start_time:.2f}s")
    
    return total_integral



# Integration bounds [-2.0, 2.0]^6 to capture Gaussian peak centered at 0.5
bounds = [[-1.,-1,-1.,-1.,0.001,0.001], [1.0,1.0,1.0,1.0,1.0,1.0]]
#tol = 1e+2

vegas_grid=np.array(integ.map.extract_grid())  
  
  
def convert_vegas_array_to_regions(vegas_grid_array, divs_per_dim=2):
    """
    Converts a raw vegas grid array (e.g. shape (6, 1001)) 
    into (num_regions, 2, dim) format for Genz-Malik adaptive integration.
    
    Parameters:
    -----------
    vegas_grid_array : np.ndarray
        The array of shape (dim, nbin+1) from integ.map.extract_grid()
    divs_per_dim : int
        Number of regions to create along each axis. 
        Total regions = divs_per_dim ** dim.
        Example: 2 divisions in 6D = 64 regions.
        
    Returns:
    --------
    np.ndarray
        Array of shape (num_regions, 2, dim)
    """
    n_dim = vegas_grid_array.shape[0]
    n_bins_available = vegas_grid_array.shape[1] - 1
    
    # 1. Select the indices for downsampling (evenly spaced bin indices)
    # This preserves the adaptive density of the original grid.
    indices = np.linspace(0, n_bins_available, divs_per_dim + 1).astype(int)
    
    # 2. Extract specific endpoints for each dimension and create intervals
    dim_intervals = []
    for d in range(n_dim):
        coords = vegas_grid_array[d][indices]
        # Create pairs of [lower, upper] for each division
        intervals = [(coords[i], coords[i+1]) for i in range(divs_per_dim)]
        dim_intervals.append(intervals)
        
    # 3. Create the Cartesian product of all dimensions to form hyper-boxes
    regions = []
    for hyperbox in itertools.product(*dim_intervals):
        # hyperbox is a sequence of dim (low, high) pairs
        low_bounds = [bound[0] for bound in hyperbox]
        high_bounds = [bound[1] for bound in hyperbox]
        
        # Format as [ [low0, low1...], [high0, high1...] ]
        regions.append([low_bounds, high_bounds])
        
    return np.array(regions)
  
vegas_conv_grid = convert_vegas_array_to_regions(vegas_grid, divs_per_dim=2)  
      
      
def convert_vegas_to_stratified_regions(integ_loc, neval_target=None):
    """
    Extracts the stratified integration boxes from a trained vegas integrator.
    This uses the full resolution of the 1001 grid nodes but avoids the 
    Cartesian explosion by following vegas's internal stratification logic.
    
    Parameters:
    -----------
    integ : vegas.Integrator
        The trained integrator (after warm-up).
    neval_target : int
        Desired number of regions. If None, uses the integrator's setting.
        
    Returns:
    --------
    np.ndarray
        Array of shape (num_regions, 2, dim)
    """
    amap = integ_loc.map
    n_dim = amap.dim
    
    # nstrat is an attribute of the Integrator, not the AdaptiveMap
    nstrat = integ_loc.nstrat
    num_regions = np.prod(nstrat)
    
    print(f"Extracting stratified grid: {nstrat} divisions per dim.")
    print(f"Generating {num_regions:,} high-resolution regions.")

    # Get the physical bin boundaries (the 1001 nodes per axis)
    raw_grid = np.array(integ.map.grid) 
    n_bins_total = raw_grid.shape[1] - 1
    
    # Prepare indices for each dimension's strata
    strata_indices = [np.arange(n) for n in nstrat]
    
    regions = []
    # Cartesian product of strata indices (NOT the 1000 bins)
    for idx_tuple in itertools.product(*strata_indices):
        low_vec = []
        high_vec = []
        for d, s_idx in enumerate(idx_tuple):
            # Map stratum index to the high-res grid node index
            # Stratum i spans from node (i * 1000/nstrat) to ((i+1) * 1000/nstrat)
            start_node = int(round(s_idx * n_bins_total / nstrat[d]))
            end_node = int(round((s_idx + 1) * n_bins_total / nstrat[d]))
            
            low_vec.append(raw_grid[d][start_node])
            high_vec.append(raw_grid[d][end_node])
            
        regions.append([low_vec, high_vec])
        
    return np.array(regions)
      
        
#print(f"Grid Shape: {my_grid.shape}")
#explicitly set seed
gvar.ranseed(99990)
integ = vegas.Integrator(mx.array([[-1, 1],[-1,1],[-1,1],[-1,1],[0.0001,1],[0.0001,1]],dtype=mx.float32))
# STAGE 1: Warm-up
res_warmup = integ(model, nitn=10, neval=50000)
#vegas_grid=np.array(integ.map.extract_grid()) 

vegas_conv_grid2=convert_vegas_to_stratified_regions(integ)

result = adaptive_integration_with_grid(initial_grid=vegas_conv_grid2, bounds=bounds,tol=1e+2)
    
log_evidence = mymax + mx.log(result)    
print(f"log_evidence={log_evidence}\n")



```


```{python genz_a0,include=FALSE,eval=TRUE}
import mlx.core as mx
import numpy as np
import heapq
import itertools
import time
import mlx 

# 1. GPU-Optimized Integrand

@mx.compile
def integrand_gpu2(pts):
    theta=mx.transpose(pts)
    #print(f"shape1={x.shape}\n")
    """
    as above but we integrate out theta1-theta5 for fixed theta0
    
    """
    # theta0idx=NA # Not needed as pass in function as not marginalizing this (index for a0 transformed)
    theta1_idx=0 # index for a1 - logit treatment effect transformed
    theta2_idx=1 # index for mu transformed
    theta3_idx=2 # index for m1 transformed
    theta4_idx=3 # index for sigma0 transformed
    theta5_idx=4 # index for sigma1 transformed
    
    theta1=mx.clip(theta[theta1_idx,:],-0.99999, 0.99999)
    theta2=mx.clip(theta[theta2_idx,:],-0.99999, 0.99999)
    theta3=mx.clip(theta[theta3_idx,:],-0.99999, 0.99999)
    theta4=mx.clip(theta[theta4_idx,:],0.00001, 0.99999)
    theta5=mx.clip(theta[theta5_idx,:],0.00001, 0.99999)
    
    
    jacobianL = (  mx.log(1+mx.power(theta1,2)) - 2.0*mx.log(1-mx.power(theta1,2))
                 + mx.log(1+mx.power(theta2,2)) - 2.0*mx.log(1-mx.power(theta2,2))
                 + mx.log(1+mx.power(theta3,2)) - 2.0*mx.log(1-mx.power(theta3,2))
                 + mx.log(1+mx.power(theta4,2)) - 2.0*mx.log(1-mx.power(theta4,2))
                 + mx.log(1+mx.power(theta5,2)) - 2.0*mx.log(1-mx.power(theta5,2))
                 )
    #print(f"jacobian={jacobianL}\n")             

    # a0 is fixed and passed into likelihood
    a1=theta1/(1-mx.power(theta1,2))
    mu0=theta2/(1-mx.power(theta2,2))
    mu1=theta3/(1-mx.power(theta3,2))
    sigma0=theta4/(1-mx.power(theta4,2))
    sigma1=theta5/(1-mx.power(theta5,2))
    #print(f"a0={a0} a1={a1} mu0={mu0} mu1={mu1} sigma0={sigma0} sigma1={sigma1}\n")
    
    eta=a0+a1*T # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - mx.log(1+mx.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = mx.sum(y*eta - mx.log(1+mx.exp(eta)),axis=0) #
    #print(f"logL={logL}\n")
    # now add the priors
    # simple N(0,sd=10)
    prior_a0 = mx_norm_logpdf(a0,loc=mu0,scale=sigma0)
    prior_a1 = mx_norm_logpdf(a1,loc=mu1,scale=sigma1)
    prior_mu0 = mx_norm_logpdf(mu0,loc=0.,scale=2.5)
    prior_mu1 = mx_norm_logpdf(mu1,loc=0.,scale=2.5)
    prior_sigma0 = mx_half_normal_logpdf(sigma0, 2.5)
    prior_sigma1 = mx_half_normal_logpdf(sigma1, 2.5)
    
    logDens = logL + prior_a0 + prior_a1 + prior_mu0 + prior_mu1 + prior_sigma0 + prior_sigma1
    #print(f"logDens={logDens}\n")
    #print(f"mymax2={mymax} a0={a0}\n")
    return mx.exp(logDens + jacobianL - mymax) 

@mx.compile
def evaluate_batch_gpu(m, h):
    """
    Fully vectorized evaluation of B regions on GPU.
    """
    # Broadcasting: (B, 1, n) + (1, N, n) * (B, 1, n) -> (B, N, n)
    pts = m[:, None, :] + REL_PTS[None, :, :] * h[:, None, :]
    
    # Flatten and evaluate
    vals = integrand_gpu2(mx.reshape(pts, (-1, n)))
    vals = mx.reshape(vals, (-1, num_points))
    
    # Calculate volumes (B,)
    volumes = mx.prod(2.0 * h, axis=1)
    
    # Calculate Degree 7 and Degree 5 integrals
    i7 = mx.sum(vals * WEIGHTS_7, axis=1) * volumes
    i5 = mx.sum(vals * WEIGHTS_5, axis=1) * volumes
    errs = mx.abs(i7 - i5)
    
    return i7, errs


#start = jnp.array([-1,-1,-1,-1,0.0001,0.0001],dtype=jnp.float32) # a0 a1
#stop = jnp.array([1,1,1,1,1,1],dtype=jnp.float32)   # a0 a1
#searchPts=mx.array(np.array(jnp.transpose(jnp.linspace(start, stop, 1000))), dtype=mx.float32)

#mx.max(integrand_gpu(mx.transpose(searchPts)))


# 2. Genz-Malik Degree 7 Rule for n=6
n = 5
lam2 = np.sqrt(9/70)
lam4 = np.sqrt(9/10)
lam5 = np.sqrt(9/19)

nodes = []
nodes.append([0.0] * n) # Center (1 point)

for i in range(n): # Axial lam2 (2n points)
    p1, p2 = [0.0] * n, [0.0] * n
    p1[i], p2[i] = lam2, -lam2
    nodes.append(p1); nodes.append(p2)

for i in range(n): # Axial lam4 (2n points)
    p1, p2 = [0.0] * n, [0.0] * n
    p1[i], p2[i] = lam4, -lam4
    nodes.append(p1); nodes.append(p2)

for i, j in itertools.combinations(range(n), 2): # Cross lam4 (2n(n-1) points)
    for si, sj in [(1,1), (1,-1), (-1,1), (-1,-1)]:
        p = [0.0] * n
        p[i], p[j] = si * lam4, sj * lam4
        nodes.append(p)

for signs in itertools.product([1.0, -1.0], repeat=n): # Diagonal lam5 (2^n points)
    nodes.append([s * lam5 for s in signs])

REL_PTS = mx.array(nodes) 
num_points = REL_PTS.shape[0]
print(f"Generated {num_points} points for {n}D Genz-Malik rule.")

# Pre-calculate Weights as MLX arrays
w1 = (12824 - 9120*n + 400*n**2) / 19683
w2, w4 = 980 / 6561, 200 / 19683
w3, w5 = (1820 - 400*n) / 19683, 6859 / (19683 * 2**n)
WEIGHTS_7 = mx.array([w1] + [w2]*(2*n) + [w3]*(2*n) + [w4]*(2*n*(n-1)) + [w5]*(2**n))

we1 = (729 - 950*n + 50*n**2) / 729
we2, we4 = 245 / 486, 25 / 729
we3, we5 = (265 - 100*n) / 1458, 0.0
WEIGHTS_5 = mx.array([we1] + [we2]*(2*n) + [we3]*(2*n) + [we4]*(2*n*(n-1)) + [we5]*(2**n))


@mx.compile
def compute_log_lik2(theta, y,T,a0):
    #print(f"shape1={x.shape}\n")
    """
    as above but we integrate out theta1-theta5 for fixed theta0
    
    """
    # theta0idx=NA # Not needed as pass in function as not marginalizing this (index for a0 transformed)
    theta1_idx=0 # index for a1 - logit treatment effect transformed
    theta2_idx=1 # index for mu transformed
    theta3_idx=2 # index for m1 transformed
    theta4_idx=3 # index for sigma0 transformed
    theta5_idx=4 # index for sigma1 transformed
    
    theta1=mx.clip(theta[theta1_idx,:],-0.99999, 0.99999)
    theta2=mx.clip(theta[theta2_idx,:],-0.99999, 0.99999)
    theta3=mx.clip(theta[theta3_idx,:],-0.99999, 0.99999)
    theta4=mx.clip(theta[theta4_idx,:],0.00001, 0.99999)
    theta5=mx.clip(theta[theta5_idx,:],0.00001, 0.99999)
    
    
    jacobianL = (  mx.log(1+mx.power(theta1,2)) - 2.0*mx.log(1-mx.power(theta1,2))
                 + mx.log(1+mx.power(theta2,2)) - 2.0*mx.log(1-mx.power(theta2,2))
                 + mx.log(1+mx.power(theta3,2)) - 2.0*mx.log(1-mx.power(theta3,2))
                 + mx.log(1+mx.power(theta4,2)) - 2.0*mx.log(1-mx.power(theta4,2))
                 + mx.log(1+mx.power(theta5,2)) - 2.0*mx.log(1-mx.power(theta5,2))
                 )
    #print(f"jacobian={jacobianL}\n")             

    # a0 is fixed and passed into likelihood
    a1=theta1/(1-mx.power(theta1,2))
    mu0=theta2/(1-mx.power(theta2,2))
    mu1=theta3/(1-mx.power(theta3,2))
    sigma0=theta4/(1-mx.power(theta4,2))
    sigma1=theta5/(1-mx.power(theta5,2))
    #print(f"a0={a0} a1={a1} mu0={mu0} mu1={mu1} sigma0={sigma0} sigma1={sigma1}\n")
    
    eta=a0+a1*T # (10,3) where T_vec = (3,) and (10,1) broadcast to (10,3)  
    # y_data*eta - mx.log(1+mx.exp(eta)) this is (10,3) - want col sums, so collapse over rows
    logL = mx.sum(y*eta - mx.log(1+mx.exp(eta)),axis=0) #
    #print(f"logL={logL}\n")
    # now add the priors
    # simple N(0,sd=10)
    prior_a0 = mx_norm_logpdf(a0,loc=mu0,scale=sigma0)
    prior_a1 = mx_norm_logpdf(a1,loc=mu1,scale=sigma1)
    prior_mu0 = mx_norm_logpdf(mu0,loc=0.,scale=2.5)
    prior_mu1 = mx_norm_logpdf(mu1,loc=0.,scale=2.5)
    prior_sigma0 = mx_half_normal_logpdf(sigma0, 2.5)
    prior_sigma1 = mx_half_normal_logpdf(sigma1, 2.5)
    
    logDens = logL + prior_a0 + prior_a1 + prior_mu0 + prior_mu1 + prior_sigma0 + prior_sigma1
    #print(f"logDens={logDens}\n")
    return (logDens + jacobianL) # this will be (3,)


# MARGINAL a0 
# Setup Vegas with a Shifting Logic
@vegas.rbatchintegrand
class StableLikelihood2:
    def __init__(self, y, T,a0):
        self.y = y
        self.T = T
        self.l_max = 0.0 # Will be updated after warm-up
        self.a0 = a0 # random variable which we want to compute density value for

    def __call__(self, theta):
        theta_mx = mx.array(theta)
        log_lik = compute_log_lik2(theta_mx, self.y,self.T, self.a0)
        
        # Shift by the maximum found so far
        # This keeps the result near 1.0 instead of 1e-300
        ## return np.array(jnp.exp(log_lik - self.l_max)) # needt this is using x64 - vegas expects np
        return mx.exp(log_lik - self.l_max)


# Integration bounds [-2.0, 2.0]^6 to capture Gaussian peak centered at 0.5
bounds = [[-1.,-1.,-1.,0.001,0.001], [1.0,1.0,1.0,1.0,1.0]]
#tol = 1e+2

#vegas_grid=np.array(integ.map.extract_grid())  
  
#print(f"Grid Shape: {my_grid.shape}")
#explicitly set seed
gvar.ranseed(99991)

a0=-1.0
start = jnp.array([-1,-1,-1,0.0001,0.0001],dtype=jnp.float32) # a0 a1
stop = jnp.array([1,1,1,1,1],dtype=jnp.float32)   # a0 a1
searchPts=mx.array(np.array(jnp.transpose(jnp.linspace(start, stop, 1000))), dtype=mx.float32)
model2 = StableLikelihood2(y=y_data,T=T_vec,a0=a0)
model2.l_max=mx.max(compute_log_lik2(searchPts, model2.y,model2.T,model2.a0))
mymax=model2.l_max

integ2 = vegas.Integrator(mx.array([[-1, 1],[-1,1],[-1,1],[0.0001,1],[0.0001,1]],dtype=mx.float32))
# STAGE 1: Warm-up
res_warmup = integ2(model2, nitn=10, neval=50000)
#vegas_grid=np.array(integ.map.extract_grid()) 

vegas_conv_grid2=convert_vegas_to_stratified_regions(integ2)

#result = adaptive_integration_with_grid(bounds=bounds,tol=1e-02)
# this also works
result = adaptive_integration_with_grid(initial_grid=vegas_conv_grid2,tol=1e-02)
    
marg_a0 = mymax + mx.log(result)   # model.l_max + mx.log(gvar.mean(result))
print(f"log_evidence={log_evidence}\n")

myprobden=mx.array(mx.exp(marg_a0-log_evidence))
print(f"den value at a0={a0} = {myprobden}\n")

```



